### 2016-07-05 introduction.tex

Recent results on reasoning about application invariants in the
presence of weak \emph{consistency}~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that usually manifests on replicated
data stores, where operations are concurrently executed against
different replicas, resulting in an order of execution inconsistent
with their serial order. The operations, nonetheless, are atomic and
fully isolated, and each operation is required to locally preserve
application invariants. In contrast, weak isolation is a property of
transactions, which are sets of atomic operations. Weak isolation
manifests when successive atomic operations in a transaction witness
different contemporary states (or different replicas) of the database
which, although consistent individually, may not be obviously
reconciled into a consistent global view. Furthermore, operations may
temporarily break invariants as long as transaction as a whole
preserves them.  Intuitively, the latter problem reduces to the former
if all transactions contain a single operation. Furthermore, weak
isolation can exist independent of weak consistency, as evident from
the presence of weakly isolated transactions on sequentially
consistent relational databases; such systems existed long before the
advent of weakly consistent NoSQL stores.  Thus, frameworks for
reasoning about weak isolation will necessarily have to generalize the
reasoning frameworks developed for weak consistency in new and
important ways.

@@

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on reasoning about programs under
weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
model does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency, and generates executions that are guaranteed to satisfy
these specifications. The result is a flexible system model that is
general enough to incorporate the semantics of a range of isolation
levels on a variety of stores (\eg the sequentially consistent store
of~\cite{adyaphd}, or the causally consistent store
of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
a set of proof rules that demonstrate that a program with a given
selection of isolation levels for its transactions preserves its
invariants when executed on a store equipped with certain consistency
guarantees (\eg causal consistency).

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly-consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

\noindent Our results provide the first (to the best of our
knowledge) mechanism that precisely relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

>>>>

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that usually manifests on replicated
data stores, where operations are concurrently executed against
different replicas, resulting in an order of execution inconsistent
with their serial order. The operations, nonetheless, are atomic and
fully isolated, and each operation is required to locally preserve
application invariants. In contrast, weak isolation is a property of
transactions, which are sets of atomic operations. Weak isolation
manifests when successive atomic operations in a transaction witness
different contemporary states (or different replicas) of the database
which, although consistent individually, may not be obviously
reconciled into a consistent global view. Furthermore, operations may
temporarily break invariants as long as transaction as a whole
preserves them.  Intuitively, the latter problem reduces to the former
if all transactions contain a single operation. Furthermore, weak
isolation can exist independent of weak consistency, as evident from
the presence of weakly-isolated transactions on sequentially
consistent relational databases; such systems existed long before the
advent of weakly consistent NoSQL stores.  Thus, frameworks for
reasoning about weak isolation will necessarily have to generalize the
reasoning frameworks developed for weak consistency in new and
important ways.

@@

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly-isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on reasoning about programs under
weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
model does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency, and generates executions that are guaranteed to satisfy
these specifications. The result is a flexible system model that is
general enough to incorporate the semantics of a range of isolation
levels on a variety of stores (\eg the sequentially consistent store
of~\cite{adyaphd}, or the causally consistent store
of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
a set of proof rules that demonstrate that a program with a given
selection of isolation levels for its transactions preserves its
invariants when executed on a store equipped with certain consistency
guarantees (\eg causal consistency).

The paper makes the following contributions:

\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We present a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly-consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

\noindent Our results provide the first (to the best of our knowledge)
mechanism that precisely and uniformly relates high-level program
invariants to low-level weak isolation and store consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

### 2016-07-05 opsem.tex

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behavior is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behavior as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, hence lets us capture finer
properties than $\hbZ$. We therefore let $\visZ$ replace $\hbZ$ in our
discussion henceforth. The session order relation captures the
sequential order of operations within a transaction. In particular, it
relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.

>>>>

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behavior is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behavior as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

@@

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, and hence lets us capture
finer-grained isolation properties than $\hbZ$, which we leverage in
our development.  The session order relation captures the sequential
order of operations within a transaction. In particular, it relates
two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.
